---
title: "16S Data Processing Training"
author: Katie McCauley from the Bioinformatics and Computational Biology Branch (BCBB)
format: 
  html:
    toc: true
    toc-location: left
    number-sections: true
    embed-resources: true
    code-overflow: wrap
    theme: yeti
editor: visual
---

## Introduction

Thank you for joining us for this 16S rRNA microbiome data processing tutorial! Before we jump in, let's start by going over our learning objectives for the technical part.

-   Identify and Remove Primers and Adapters (if present)

-   Check the Quality of Your Microbiome Data

-   Develop Amplicon Sequence Variants (ASVs) using Divisive Amplicon Denoising Algorithm (DADA2)

-   Assign taxonomy to the species level

-   Make a phylogenetic tree

-   Collate this data in a `phyloseq` object for later analysis

This tutorial is designed to introduce you to the nuts and bolts of turning sequencing data into analyzable counts with emphasis on. You may find that after you finish this tutorial, you start processing your data through other pipelines like Nephele[^1], but hopefully some of the concepts we cover today will guide how you approach and explore your data.

[^1]: Disclaimer: The tools, functions and options that we will use today are not precisely the ones used by Nephele

## Find and Organize our Data

First things first: we need to organize our data! If you didn't bring your own data, you will find the files that we work with today under `raw_data`, and we will use this location throughout the tutorial today.

```{r}
library(dada2)

setwd("~/Documents/training/16S-data-processing/")
read_indicator <- "_R1"
all_fastqs <- list.files("raw_data", full.names = T)
all_fastqs
r1_fastqs <- all_fastqs[grepl(read_indicator, all_fastqs)]
r2_fastqs <- all_fastqs[!grepl(read_indicator, all_fastqs)]
r1_fastqs
r2_fastqs
```

## Identifying Primers and Adapters

For the purposes of our discussion, I'm going to use R to ask if there are any primers in our sequencing data. Our data was derived from a subset of samples sequenced in this manuscript (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6129204/). It uses the V4 region of the 16S rRNA gene, and uses 515F and 806R primers. You can get the primer sequences from the Earth Microbiome Project's website (https://earthmicrobiome.org/protocols-and-standards/16s/). You may need to obtain different sequences for your primer strategy, but the overall code to search for our (known) primers remains the same.

```{r}
library(ShortRead)
library(Biostrings)
FWD <- "GTGCCAGCMGCCGCGGTAA"
REV <- "GGACTACHVGGGTWTCTAAT"

primerHits <- function(primer, fn) {
  nhits <- vcountPattern(primer, sread(readFastq(fn)), fixed = FALSE)
  return(sum(nhits > 0))
}
sapply(FWD, primerHits, r1_fastqs)
sapply(REV, primerHits, r2_fastqs)
nchar(FWD)
nchar(REV)
```

We indeed have primer sequences! For our purposes, we'll use DADA2's filterAndTrim function to help us remove them before analysis.

## Quality Filtering and Trimming

First, let's use the `dada2` package's ability to tell us a little bit more about the quality profile of our samples. You may have already seen this for your data if you used Nephele's QC pipeline, but this allows us to take a look at our data again before we apply filtering and trimming.

```{r}
plotQualityProfile(r1_fastqs)
plotQualityProfile(r2_fastqs)

```

Given this information, we will use the `filterAndTrim` function to clean our data. Before we start, we need somewhere for our filtered reads to go, so we will make those paths here. Keep in mind that I'm performing very simple string modification, but you may need to do something more complex for your data and samples.

```{r}
r1_filt <- gsub("raw_data", "filtered", r1_fastqs)
r2_filt <- gsub("raw_data", "filtered", r2_fastqs)
## In my case, the filtered directory could exist before I make the document -- this `unlink` function just makes sure it's not there when I run filterAndTrim
unlink("filtered/", recursive = T)
out <- filterAndTrim(r1_fastqs, r1_filt, r2_fastqs, r2_filt, trimLeft=c(19,20), truncLen=c(200, 140), maxN=0, maxEE=2, truncQ=5, rm.phix = TRUE, compress=TRUE, multithread = F, verbose=T)
out
```

## Denoise Reads

```{r}
r1_err <- learnErrors(r1_filt, multithread=F, verbose=T)
r2_err <- learnErrors(r2_filt, multithread=F, verbose=T)

plotErrors(r1_err, nominalQ = T)
plotErrors(r2_err, nominalQ = T)

r1_dada <- dada(r1_filt, r1_err)
r2_dada <- dada(r2_filt, r2_err)
```

## Merge Reads

```{r}
merged_reads <- mergePairs(r1_dada, r1_filt, r2_dada, r2_filt, verbose=TRUE, minOverlap = 25)
head(merged_reads[[2]])
```

## Make Sequence Table

```{r}
seq_table <- makeSequenceTable(merged_reads)
dim(seq_table)
```

From this information, we can see that there are three rows for each of the three samples and there are 3,628 columns for each of the sequence variants.

We can use basic R functions to explore the dataset, much like we would any table of counts.

Below is the distribution of the sequence lengths. Notice that most sequences are 253 bases long, which is how long the V4 region is. If you're unsure if you have primers, this can be a good point to sanity check.

```{r}
table(nchar(getSequences(seq_table)))
```

::: {.callout-tip appearance="simple"}
## Filter for sequence length!

If you look at the table and find that you have sequences that are much larger or smaller than your target amplicon, you can remove through basic R commands like: `filt_seq_table <- seq_table[, nchar(getSequences(seq_table)) >= median(nchar(getSequences(seq_table))]`
:::

## Remove Chimeras

```{r}
chimera_check <- removeBimeraDenovo(seq_table, verbose=T)
dim(chimera_check)
```

::: callout-important
## Workflow Checkpoint!

Let's take a look at how our data looks so far in a nice, convenient table!

```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(r1_dada, getN), sapply(r2_dada, getN), sapply(merged_reads, getN), rowSums(chimera_check))
colnames(track) <- c("input","filtered","denoised_r1","denoised_r2","merged","nonchimeric")
track
```

As you can see, there was a slight loss of reads during chimera checking for these samples. For now, it isn't enough to go back and double-check our parameters, but if you find that you lose about half of our reads due to chimeras, you may need to revisit the chimera method. For instance, we actually used the "pooled" chimera checking method because we lost many more sequences with the "consensus" chimera checking method, but for other data, you may need to look further back.
:::

## Assign Taxonomy

We now have `{r} ncol(chimera_check)` sequence variants and we want to determine what bacteria they are assigned to. In order to do this, we need to download specially-formatted databases (you can find several different databases [here](https://benjjneb.github.io/dada2/training.html), some of which are contributed by outside teams). For our purposes, we'll download the [SILVA database](https://zenodo.org/record/4587955)\^\[Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies J, Glöckner FO (2013) The SILVA ribosomal RNA gene database project: improved data processing and web-based tools. Nucl. Acids Res. 41 (D1): D590-D596.\]. You'll need all three files and we'll install them to a folder called `databases` that is in our working directory.

```{r}
taxa <- assignTaxonomy(chimera_check, "databases/silva_nr99_v138.1_train_set.fa.gz", multithread=F, verbose=T, minBoot = 80)
taxa_print <- taxa
rownames(taxa_print) <- NULL
head(taxa_print,10)
```

## Make Phylogenetic Tree

Let's arrange our unique sequences into a phylogenetic tree so that we know how similar they are to each other. This tree can be used for statistical questions or visualizations later.

```{r}
library(DECIPHER)
aligned <- DECIPHER::AlignSeqs(DNAStringSet(getSequences(chimera_check)))
```

## Generate a Phyloseq Object
